---
title: "README"
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


The **estimateR** package provides tools to estimate the effective reproductive number through time from delayed and indirect observations of infection events.
This package is currently in a beta version. 

## Installation

First, make sure you have installed the **devtools** package locally.
If not, run:
```
install.packages("devtools")
``` 
in RStudio or other.

Then run:
```
library(devtools)
install_github("covid-19-Re/estimateR")
```

## Documentation

The full documentation is available **[here](https://covid-19-re.github.io/estimateR/)**.

## Quick example

We demonstrate below a basic use of the **estimateR** package. Check out the *estimateR* vignette and additional vignettes for more details.

We start with arbitrary incidence data representing daily counts of case confirmations for a disease of interest.
This incidence data counts the number of people getting tested positive for a particular infectious disease X on day *T*.
With `estimate_Re_from_noisy_delayed_incidence()`, we compute the reproductive number through time from this incidence data
and specifications of the delays between infection events and case confirmations as well as the serial interval.

```{r}
library(estimateR)
date_first_data_point <- as.Date("2020-02-24")
toy_incidence_data <- c(4,9,19,14,36,16,39,27,46,
                          77,78,113,102,134,165,183,
                          219,247,266,308,304,324,346,
                          348,331,311,267,288,254,239)


## The numbers below are part of the user input, 
## they need to be adapted to the particular disease studied.

# Incubation period - gamma distribution parameters
shape_incubation = 3.2 
scale_incubation = 1.3
incubation <- list(name="gamma", shape = shape_incubation, scale = scale_incubation)

# Delay from onset of symptoms to case observation - gamma distribution parameters
shape_onset_to_report = 2.7
scale_onset_to_report = 1.6
onset_to_report <- list(name="gamma", shape = shape_onset_to_report, scale = scale_onset_to_report)

# We specify these parameters in the same unit as the time steps in the original observation data.
# For instance, if the original data represents daily reports,
# the parameters below must be specified in days (this is the case in this toy example).
mean_serial_interval = 4.8
std_serial_interval = 2.3

# The estimation window corresponds to the size of the sliding window used in EpiEstim.
# See help(estimate_Re) for additional details.
# Here, it is set to three days.
estimation_window = 3 

## End of user input
toy_estimates <- estimate_Re_from_noisy_delayed_incidence(toy_incidence_data,
  smoothing_method = "LOESS",
  deconvolution_method = "Richardson-Lucy delay distribution",
  estimation_method = "EpiEstim sliding window",
  delay = list(incubation, onset_to_report),
  estimation_window = estimation_window,
  mean_serial_interval = mean_serial_interval,
  std_serial_interval  = std_serial_interval,
  output_Re_only = FALSE,
  ref_date = date_first_data_point,
  time_step = "day"
)

tail(toy_estimates, n = 20)
```



